---
title: "Literature Review"
bibliography: ../01_Input/references.bib
csl: ../01_Input/agu.csl
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Table of Metablic Processes 

Last compiled on *`r format.POSIXct(Sys.time(), format = "%e %B %Y %H:%M")`*

## Table 1


```{r results = "asis"}

tabledata <- rbind.data.frame(
  list(
    MetabolicProcess = "Aerobic respiration",
    Reaction = "$\\mathrm{\\frac{1}{2} CH_{3}COOH}$", 
    FreeEnergy = 437
  )
)
names(tabledata) <- c(
  "Metabolic Process",
  "Reaction",
  "Free Energy"
)

cat(
  print(
    knitr::kable(
      tabledata,
      col.names = c(
        "Metabolic Process",
        "Reaction",
        "Free Energy"
      )
    )
  ),
  "\n\n"
)



```

`r knitr::kable(tabledata)`


## Table 2

```{r results = "asis"}

tabledata2 <- data.frame(
  MetabolicProcess = c("Aerobic respiration",
                       "Denitification:",
                       "- *Nitrate reduction undefined*",
                       "- *Nitrate reduction*",
                       "- *Nitrus oxide reduction*",
                       "Sulfate reduction", 
                       "Methanogenesis",
                       "Nitrification:",
                       "*Ammonium oxidation*",
                       "*Nitrate oxidation*",
                       "Methantropy"),
  Reaction = c("$\\mathrm{\\frac{1}{2} CH_{3}COOH + O_{2} \\longleftrightarrow CO_{2} + H_{2}O}$&nbsp;", 
               "$\\mathrm{\\frac{1}{2} CH_{3}COOH + \\frac{4}{5}NO_{3}{}^{-} + \\frac{4}{5}H^{+} \\longleftrightarrow CO_{2} + \\frac{2}{5}N_{2} + \\frac{7}{5}H_{2}O}$&nbsp;",
               "$~~\\mathrm{\\frac{1}{2} CH_{3}COOH + {2}NO_{3}{}^{-} \\longleftrightarrow CO_{2} + {2}NO_{2}{}^{-} + H_{2}O}$&nbsp;",
               "$~~\\mathrm{\\frac{1}{2} CH_{3}COOH + {2}NO_{2}{}^{-} + {2}H^{+} \\longleftrightarrow CO_{2} + N_{2}O + 2H_{2}O}$&nbsp;",
               "$~~\\mathrm{\\frac{1}{2} CH_{3}COOH + {2}N_{2}O \\longleftrightarrow CO_{2} + 2N_{2} + H_{2}O}$&nbsp;",
               "$\\mathrm{\\frac{1}{2} CH_{3}COOH + \\frac{1}{2} SO_{4}{}^{-2} + \\frac{1}{2} H^{+} \\longleftrightarrow \\frac{1}{2}HS^{-} + CO_{2} + H_{2}O}$&nbsp;",
               "$\\mathrm{\\frac{1}{2} CH_{3}COOH \\longleftrightarrow \\frac{1}{2} CO_{2} + \\frac{1}{2} CH_{4}}$&nbsp;",
               "$\\mathrm{O_{2} + \\frac{1}{2}NH_{4}{}^{+} \\longleftrightarrow \\frac{1}{2}NO_{3}{}^{-} + H^{+} + \\frac{1}{2}H_{2}O}$&nbsp;",
               "$~~\\mathrm{O_{2} + \\frac{2}{3}NH_{4}{}^{+} \\longleftrightarrow \\frac{2}{3}NO_{2}{}^{-} + \\frac{4}{3}H^{+} + \\frac{2}{3}H_{2}O}$&nbsp;",
               "$~~\\mathrm{O_{2} + 2NO_{2}{}^{-} \\longleftrightarrow {2}NO_{3}{}^{-}}$&nbsp;",
               "$\\mathrm{O_{2} + \\frac{1}{2}CH_{4} \\longleftrightarrow \\frac{1}{2}CO_{2} + H_{2}O}$&nbsp;"),
  FreeEnergy = c(437, 410, 288, 415, 645, 38, 28, 174, 183, 148, 409)
)


names(tabledata2) <- c(
  "Metabolic Process",
  "Reaction",
  "$E_S = -\\Delta G_{0}^{'} \\mathrm{(kJ ~mol^{-1})}$"
)


cat(
  print(
    knitr::kable(
      tabledata2,
      col.names = c(
        "Metabolic Process",
        "Reaction",
        "$E_S = -\\Delta G_{0}^{'} \\mathrm{(kJ ~mol^{-1})}$"
      )
    )
  ),
  "\n\n"
)


```

`r knitr::kable(tabledata2)`


<table>
  <tr>
    <th>Metabolic Process</th>
    <th>Reaction</th>
    <th>$E_S = -\Delta G_{0}^{'} \mathrm{(kJ ~mol^{-1})}$</th>
  </tr>
</table>

# N-DAMO

*@zhang_nitrite-dependent_2024*

### History

earliest mentions, notable papers, etc

*@zhu_combined_2011, @luesken_simultaneous_2011*

### Where N-DAMO fits

co-occurances, influences, etc

# Modeling metabolic processes

strategies, notable papers, other instances of modeling

*@hall_linking_2011, @he_modelling_2013, @payn_generalized_2014, @reinhold_constraint-based_2019*

\n\n

# Annotated Bibliography

1. **@budinich_multi-objective_2017**
  - constraint-based modelling to build predictive models
  - multi-objective flux balance analysis (MO-FBA) and multi-objective flux variability analysis (MO-FVA)
  
2. **@ettwig_nitrite-driven_2010**
  - one of earliest n-damo papers
  - presents evidence of new pathway producing oxygen
      - couples anaerobic oxidation of methane with nitrate/nitrite reduction
  - assembles complete genome of dominate bacterium, M.oxyfera
  - results explain biochemical mechanism of poorly understood freshwater methane sink
  
3. **@gomez-gallego_n-damo_2022**
  - annotations
4. **@hall_linking_2011**
  - Mini-review of applications of ecological stoichiometry
  - synthesizes both theoretical frameworks and empirical studies to demonstrate how understanding the stoichiometric relationships between microbes and their environment can explain nutrient cycling, energy flow, and ecosystem stability
5. **@he_modelling_2013**
  - develops and evaluates kinetic model to describe N-DAMO
      - based on Monod type kinetics and diffusion-reaction model 
  - key kinetic parameters for n-damo bacteria identified
  - methane not a limiting factor for n-damo process
6. **@helton_thermodynamic_2015**
  - Asking whether patterns of DOC, nitrite, and ammonium concentrations align w/ conceptual model of stoichiometric and thermodynamic controls. 
  - results suggest that the relationships between both nitrate and DOC and nitrate and ammonium arise from the interaction of microbial biomass stoichiometry and the availability of terminal electron acceptors.
      - to understand biogeochemical patterns, need conceptual and qualitative models integrating thermodynamic, stoichiometric, and evolutionary controls on biogeochemical cycling 
7. **@hu_evidence_2014**
  - obtained evidence for n-damo in 3 freshwater wetlands in China
  - determined species presence and abundance of group A of M.oxyfera-like bacteria
      - previously demonstrated as dominant bacteria in n-damo process
  - estimated methane sink through n-damo process in wetlands
      - 4.1-6.1 Tg average each year under anaerobic conditions
8. **@hu_optimization_2019**
  - focused on improving n-damo process performance through enhancing methane availability by investigating:
      - immobilized biologically activated carbon (IC)
      - high static pressure (HP)
  - both improved activity of n-damo bacteria
      - highest methane absorption + n-damo bacteria abundance in combination IC-HP
9. **@jessup_big_2004**
  - overviews studies addressing ecological questions using lab-based microbial model systems
  - states that simplicity of the model systems is what makes them powerful tools, although there are practical limitations to models
10. **@legierse_nitrate-dependent_2023**
  - evaluates n-damo potential for decreasing methane release, removing excess nitrate from agricultural drainage ditches
  - significant decrease in methane.nitrate seen only in brackish peatland site, not freshwater sediments
      - n-damo bacteria likely out-competed by heterotrophic denitrifiers 
  - demonstrates that effect of n-damo bioaugmentation is system specific
11. **@luesken_simultaneous_2011**
  - tested feasibility of coexistence of anammox and n-damo bacteria
  - coculture with equal amounts established after 161 days
      - converted nitrite at rate of 17.2 mmol per day
  - early paper discussing n-damo and applications
12. **@nie_operation_2020**
  - investigated interactions between anammox in membrane biofilm reactor (MBfR)
  - results provide promising operational strategies for high-rate of nitrogen removal in landfill leachate through integration of n-damo and anammox process
13. **@payn_generalized_2014**
  - developed model of aquatic microbial metabolism and growth
      - simulates influence of multiple possible metabolic reactions on biogeochemistry of a system simultaneously
  - contributes to generalized odeling of microbial metabolism
14. **@prosser_role_2007**
  - argues that application of theory is required most in microbial ecology yet severely lacking in the field
  - describes areas of microbial ecology in which theory/theoretical framework can and should be applied
15. **@reinhold_constraint-based_2019**
  - Presents new method for simulating multiple linked biogeochemical cycles
  - GANGSTA - Generalized Algorithm for Nutrient Growth, Stoichiometric, and Thermodynamic Analysis
  - Approach used in paper allows for automatic creation of models of complex biochemical interactions
16. **@ricciuto_integrative_2021**
  - Expanded current earth energy model to include methods of predicting methane dynamics
  - Found that with optimized model parameters, model accurately predicts methane fluxes and vertical profiles
  - Provides framework for building and developing model
17. **@riley_barriers_2011**
  - Developed and tested new methane biogeochemical model within Community Earth Systems Model
  - Model able to represent and predict methane processes but retains high level of uncertainty
18. **@schneider_complexity_1994**
  - proposes thermodynamic mdel for ecosystem development
  - recommends methods for examining and modeling ecosystems
19. **@shen_spatial_2019**
  - Investigated competition between anammox and n-damo through examining spatial distribution in riverbed sediments
  - n-damo patterns proved to be more complex, confined, and reached a peak deeper into riverbeds
  - Results found spatial separation between processes
20. **@vallino_ecosystem_2010**
  - Investigates describing ecosystem biogeochemistry through application of the maximum entropy production (MEP) principle
  - Builds optimized metabolic ecosystem network model
  - Finds and discusses both advantages and disadvantages of MEP method 


```{r results = "asis"}

cat(
  "Last compiled on *",
  format.POSIXct(x = Sys.time(), format = "%e %B %Y %H:%M"),
  "*\n\n",
  sep = ""
)

```

# References